---
title: "Deep Neural Networks for Survival Analysis with survdnn"
subtitle: "Architecture, loss functions, and prediction in a native R torch implementation."
author: "Imad EL BADISY"
date: "2026-08-19"
format:
  pdf:
    toc: true
execute:
  warning: false
  message: false
---

Right-censored survival data $\{(t_i, \delta_i, x_i)\}_{i=1}^n$ pair an observed time $t_i = \min(T_i, C_i)$ and an event indicator $\delta_i = \mathbb{1}(T_i \le C_i)$ with a covariate vector $x_i \in \mathbb{R}^p$. The Cox proportional hazards model expresses the hazard as $\lambda(t \mid x) = \lambda_0(t)\exp(\beta^\top x)$, restricting covariate effects to a linear, time-constant log-hazard ratio. `survdnn` replaces the linear predictor $\beta^\top x$ with a multilayer perceptron $f_\theta(x)$, keeping the rest of the estimation machinery (partial likelihood, Breslow baseline, time-dependent risk) intact, and extends this replacement to non-proportional-hazards and accelerated failure time formulations.[^survdnn-cran] The package is implemented natively in R on top of `torch`, avoiding a Python backend.[^survdnn-paper]

## Architecture

For an input of dimension $p$ and hidden layer sizes $h_1, \dots, h_L$, the network is a sequential composition of blocks

$$
z^{(0)} = x, \qquad z^{(l)} = \phi\!\left(\mathrm{BN}\!\left(W^{(l)} z^{(l-1)} + b^{(l)}\right)\right), \qquad l = 1, \dots, L,
$$

where $\phi$ is one of `relu`, `leaky_relu`, `tanh`, `sigmoid`, `gelu`, `elu`, or `softplus`, $\mathrm{BN}$ is an optional batch normalization layer, and dropout with rate $p_{\text{drop}}$ is applied after each block. The output layer is a single linear map $g_\theta(z^{(L)}) = W^{(L+1)} z^{(L)} + b^{(L+1)} \in \mathbb{R}$, so the whole network is $f_\theta(x) = g_\theta(z^{(L)}(x))$. Covariates are standardized before entering the network, and the standardization constants are stored on the fitted object and reapplied at prediction time.

Interpretation of the scalar output $f_\theta(x)$ depends on the loss used to train it: a log-risk score under Cox losses, a location parameter on the log-time scale under AFT, or a time-dependent risk score under Cox-Time.

## Loss functions

### Cox partial likelihood

For a batch sorted by descending observed time, the risk set at time $t_i$ is $R(t_i) = \{j : t_j \ge t_i\}$. `survdnn` trains $f_\theta$ against the negative partial log-likelihood

$$
\ell_{\text{cox}}(\theta) = -\frac{1}{n_e}\sum_{i:\,\delta_i=1}\left(f_\theta(x_i) - \log\!\!\sum_{j \in R(t_i)}\exp f_\theta(x_j)\right),
$$

where $n_e$ is the number of events. The implementation computes the inner sum with `torch_logcumsumexp` on time-sorted linear predictors, so risk-set membership falls out of the sort order rather than an explicit double loop. An L2 variant adds $\lambda \cdot \overline{f_\theta(x)^2}$ to this loss, penalizing the magnitude of the risk score directly rather than the network weights.

### Accelerated failure time (log-normal)

Under the AFT loss, the network output $\mu_\theta(x)$ is the location parameter of a log-normal model for the event time: $\log T = c + \mu_\theta(x) + \sigma Z$, $Z \sim \mathcal{N}(0,1)$, where $c$ is a location offset fixed at the mean log event-time in the training data (stored as `aft_loc`) to keep the residual $\log t - c$ near zero. The scale $\sigma$ is a single learnable parameter, optimized jointly with $\theta$ as $\exp(\log\sigma)$ for positivity. The censored negative log-likelihood combines an exact-density term for events and a survival term for censored observations,

$$
\ell_{\text{aft}}(\theta, \sigma) = \frac{1}{n}\sum_{i=1}^n \Big[\delta_i \big(\log t_i + \log\sigma + \tfrac{1}{2}z_i^2\big) - (1-\delta_i)\log S_Z(z_i)\Big], \qquad z_i = \frac{\log t_i - c - \mu_\theta(x_i)}{\sigma},
$$

with $S_Z(z) = 1 - \Phi(z)$ the standard normal survival function, evaluated through the error function for numerical stability. This is the standard parametric censored-data likelihood, with the location term replaced by a network.

### Cox-Time

The Cox-Time loss extends the Cox partial likelihood to a time-varying log-risk function $g_\theta(t, x)$, taking time as an explicit network input rather than as a fixed offset.[^coxtime-paper] The contribution of an event at $t_i$ becomes

$$
\ell_{\text{coxtime}}(\theta) = -\frac{1}{n_e}\sum_{i:\,\delta_i=1}\left(g_\theta(t_i, x_i) - \log\!\!\sum_{j \in R(t_i)}\exp\, g_\theta(t_i, x_j)\right),
$$

which differs from the Cox loss only in that the denominator re-evaluates the network at $t_i$ for every subject at risk, rather than reusing a single risk score per subject. This makes the loss non-separable in `pred` alone: it needs the network and the full covariate matrix to build the $g_\theta(t_i, x_j)$ grid at every event time. The exported `coxtime_loss(pred, true)` reflects this directly, it raises an informative error rather than silently computing something incorrect, and the working implementation is the internal factory `survdnn__coxtime_loss_factory(net)`, which `survdnn()` uses when `loss = "coxtime"`. Risk-set membership is computed on raw observed time, while the time fed to the network is centered and scaled for training stability.

## Training

Fitting proceeds by standard gradient-based optimization: for each epoch, a forward pass computes the loss on the full training set, `backward()` accumulates gradients, and one of `adam`, `adamw`, `sgd`, `rmsprop`, or `adagrad` updates the parameters (`adam`/`adamw` default to a small weight decay of `1e-4` unless overridden). Training is full-batch rather than mini-batch, so an epoch is one optimizer step over the entire dataset. Callbacks receive the epoch index and current loss after each step and can halt training early, as with `callback_early_stopping()`. Missing values in the model variables are either dropped (`na_action = "omit"`) or treated as an error (`na_action = "fail"`) before any tensor is built.

## Prediction

`predict.survdnn()` returns one of three things, depending on `type`. `type = "lp"` returns the raw network output, interpreted according to the training loss. `type = "survival"` returns predicted survival probabilities at a set of times, and `type = "risk"` returns $1 - S(t)$ at a single time.

For Cox and Cox-L2 models, survival curves are not read directly off the network; they combine the fitted risk score with a Breslow estimate of the cumulative baseline hazard $H_0(t)$, obtained by refitting a one-covariate `coxph()` on the training linear predictors:

$$
\widehat{S}(t \mid x) = \exp\!\big(-\widehat{H}_0(t)\exp(f_\theta(x))\big).
$$

For the AFT model, survival probabilities follow directly from the log-normal form, $\widehat{S}(t\mid x) = 1 - \Phi\!\left(\frac{\log t - c - \mu_\theta(x)}{\sigma}\right)$, using the same $c$ and learned $\sigma$ stored at training time.

For Cox-Time, baseline hazard increments are estimated at each observed event time $t_k$ from the training risk sets, $d\widehat{H}_0(t_k) = dN(t_k) \big/ \sum_{j \in R(t_k)} \exp\, g_\theta(t_k, x_j)$, and cumulative hazard at a query time is obtained by summing increments at all event times up to that point, evaluated with the network at each of them: $\widehat{H}(t \mid x) = \sum_{t_k \le t} \exp\, g_\theta(t_k, x)\, d\widehat{H}_0(t_k)$. This requires one forward pass per event time per subject, which is the computational cost of allowing the risk score to vary with time.

## Evaluation

Three metrics are implemented directly against predicted survival matrices rather than delegated to an external package: the concordance index (`cindex_survmat()`), evaluated at a fixed horizon by comparing $1 - S(t^\star \mid x)$ across all admissible pairs; the IPCW-weighted Brier score (`brier()`) at a single time point, using a Kaplan-Meier estimate of the censoring distribution as weights; and the integrated Brier score (`ibs_survmat()`), a trapezoidal integral of the Brier score over a grid of times, normalized by the width of the grid. `evaluate_survdnn()` wraps these for a fitted model, and `cv_survdnn()` repeats fit-and-evaluate over stratified folds via `rsample::vfold_cv()`, stratifying on the event indicator so that fold sizes stay balanced.

## Example

A model is fit with the same formula interface as `survival::coxph()`.

```{r}
library(survdnn)
library(survival)

veteran <- survival::veteran

mod <- survdnn(
  Surv(time, status) ~ age + karno + celltype,
  data       = veteran,
  hidden     = c(16, 8),
  loss       = "cox",
  epochs     = 100,
  lr         = 1e-3,
  .seed      = 1,
  verbose    = FALSE
)

mod
```

Survival curves at a grid of times follow the same `predict()` interface used for `coxph` and `survfit` objects.

```{r}
pred <- predict(mod, newdata = veteran, type = "survival", times = c(30, 90, 180))
round(head(pred), 3)
```

## Model assessment

Discrimination and calibration are read off the same predicted matrix.

```{r}
y <- model.response(model.frame(mod$formula, veteran))

cindex_survmat(y, pred, t_star = 180)
ibs_survmat(y, pred, times = c(30, 90, 180))
```

Cross-validated performance follows the same call structure, now with the model refit per fold.

```{r}
res <- cv_survdnn(
  Surv(time, status) ~ age + karno + celltype,
  data    = veteran,
  times   = c(30, 90, 180),
  metrics = c("cindex", "ibs"),
  folds   = 3,
  .seed   = 42,
  hidden  = c(16, 8),
  epochs  = 50,
  verbose = FALSE
)

summarize_cv_survdnn(res)
```

## Practical notes

The AFT and Cox-Time losses are more expensive per epoch than plain Cox: AFT adds one learnable global parameter, and Cox-Time evaluates the network on a full $n_e \times n$ grid of (event time, subject) pairs at every step, so training cost grows with the number of events rather than staying constant. Batch normalization interacts with very small batches; since `survdnn` trains full-batch by default, this is only a concern for small datasets, where turning batch normalization off (`batch_norm = FALSE`) is often more stable than leaving it on. Dropout and weight decay are the two regularizers exposed directly; `tune_survdnn()` and `gridsearch_survdnn()` search over these together with architecture and learning-rate settings under cross-validation.

[^survdnn-cran]: `survdnn` package documentation, CRAN: <https://CRAN.R-project.org/package=survdnn>.
[^survdnn-paper]: EL BADISY, I. (2025). *SurvDNN: Survival Deep Learning Models for Tabular Data*. Submitted to The R Journal.
[^coxtime-paper]: Kvamme, H., Borgan, Ø., & Scheel, I. (2019). Time-to-event prediction with neural networks and Cox regression. *Journal of Machine Learning Research*, 20(129), 1-30.
