---
title: "Dynamic RMST from predicted survival curves with tvrmst"
date: "23-05-2026"
description: "A methodological note on dynamic restricted mean survival time as an interpretable summary of individualized survival predictions."
categories: [Survival Analysis, RMST, tvrmst]
format: html
execute:
  warning: false
  message: false
---

<ul class="menu">
  <li><a href="../index.qmd">Home</a></li>
  <li><a href="index.qmd">Notes</a></li>
</ul>

<hr>

Restricted mean survival time (RMST) summarizes a survival curve on the time scale of the outcome. For a survival function $S(t) = P(T > t)$, the RMST up to horizon $\tau$ is

$$
\mu(\tau) = \int_0^\tau S(t)\,dt.
$$

When the input is an individualized survival matrix, each row is a predicted survival curve $S_i(t_j)$ for patient $i$ at grid time $t_j$. The individual dynamic RMST curve is

$$
\widehat{\mu}_i(\tau_k) = \sum_{j=1}^{k} \frac{S_i(t_{j-1}) + S_i(t_j)}{2}(t_j - t_{j-1}),
$$

using the trapezoidal rule on the prediction grid. The population summary is then

$$
\widehat{\mu}(\tau_k) = \frac{1}{n}\sum_{i=1}^n \widehat{\mu}_i(\tau_k).
$$

For two groups, a clinically readable contrast is the RMST difference

$$
\Delta(\tau) = \widehat{\mu}_B(\tau) - \widehat{\mu}_A(\tau),
$$

which can be read as the difference in expected event-free time up to $\tau$. This is often easier to explain than a hazard ratio when proportional hazards is doubtful.[^rmst-paper]

## Example

The `tvrmst` package works with survival matrices and computes RMST over one or several horizons.[^tvrmst-cran]

```{r}
library(tvrmst)

set.seed(1)
time <- seq(0, 24, by = 1)

control <- outer(
  1:6, time,
  function(i, t) exp(-0.035 * t) * (1 - 0.015 * i)
)

treatment <- outer(
  1:6, time,
  function(i, t) exp(-0.025 * t) * (1 - 0.010 * i)
)

control_mat <- as_survmat(
  control,
  time = time,
  id = paste0("C", 1:6),
  group = rep("Control", 6)
)

treatment_mat <- as_survmat(
  treatment,
  time = time,
  id = paste0("T", 1:6),
  group = rep("Treatment", 6)
)

control_rmst <- rmst_dynamic(control_mat, tau = c(6, 12, 18, 24))
treatment_rmst <- rmst_dynamic(treatment_mat, tau = c(6, 12, 18, 24))

treatment_rmst$at_tau
```

The dynamic RMST values increase with the horizon because they accumulate survival probability over time. A group contrast is computed from the two survival matrices:

```{r}
delta <- rmst_delta(treatment_mat, control_mat, tau = c(6, 12, 18, 24))
delta$at_tau
```

In this example, the treatment survival curves stay above the control curves. The absolute RMST gain grows as the horizon increases: the estimated difference is about 0.25 months at 6 months and about 2.03 months at 24 months. The interpretation is direct: by 24 months, the treatment profile has about two additional months of expected event-free time within the restricted follow-up window.

## Practical Notes

- Report the prediction time grid and the maximum horizon.
- Avoid extrapolating beyond the time support of the fitted survival model.
- Prefer several clinically meaningful horizons when the timing of separation matters.
- Interpret RMST contrasts as descriptive summaries unless the comparison is embedded in a valid causal design.

[^tvrmst-cran]: `tvrmst` package documentation, CRAN: <https://CRAN.R-project.org/package=tvrmst>.
[^rmst-paper]: Royston, P., & Parmar, M. K. B. (2013). Restricted mean survival time: an alternative to the hazard ratio for the design and analysis of randomized trials with a time-to-event outcome. *BMC Medical Research Methodology*, 13, 152.

<hr>

<footer>
<a href="index.qmd">Back to notes</a>
</footer>
